---
title: "NCAA Men's Hockey Bracket Buffalo Sabremetrics"
author: "Nicky Mader"
date: "`r format(Sys.Date(), '%m-%d-%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo  = TRUE)
packages.list <- 
  c("openxlsx", "MASS", "tidyverse", "stringr", "ggplot2", "scales", "plotly", 
    "broom", "lubridate", "stringr", "data.table", "ggrepel", "psych")
for (p in packages.list){
  if (!p %in% installed.packages()[, "Package"]) install.packages(p)
  library(p, character.only = TRUE)
}

cn <- function(x) colnames(x)
mean.na <- function(...) mean(na.rm = TRUE, ...)
sum.na  <- function(...) sum(na.rm = TRUE, ...)
min.na  <- function(...) min(na.rm = TRUE, ...)
max.na  <- function(...) max(na.rm = TRUE, ...)
nonNA   <- function(x) sum(!is.na(x))

path <- "~/GitHub/buffalo-sabremetrics/"

# To-dos
# - Improve visualization of goal differential vs rank
# - Do some kind of noodling with the regression model

```

# Clean Data

Read in the USCHO schedule data.

```{r read and clean data}

### Read data and separate ranks for team names
d <-
  read.xlsx(xlsxFile = paste0(path, "NCAA Men's Div I Hockey - Season Game Data.xlsx"), 
            sheet = "2022-23") %>%
  select(Date = Date1, Opponent, Opp.Score, Home, Score, OT) %>%
  mutate(Date = ymd(Date),
         opp_rank  = ifelse(grepl("\\(\\d+\\)", Opponent), gsub("\\((\\d+)\\).+", "\\1", Opponent), NA) %>% as.numeric(),
         home_rank = ifelse(grepl("\\(\\d+\\)", Home),     gsub("\\((\\d+)\\).+", "\\1", Home),     NA) %>% as.numeric(),
         opp  = gsub("\\(.+\\)", "", Opponent) %>% str_trim() %>% gsub(" ", "_", x=.) %>% gsub("[\\. '-]", "", x=.),
         home = gsub("\\(.+\\)", "", Home)     %>% str_trim() %>% gsub(" ", "_", x=.) %>% gsub("[\\. '-]", "", x=.)) %>%
  select(-Opponent, -Home)
teams <- with(d, c(opp, home)) %>% unique() %>% sort()
teams
```

```{r develop last rank info, eval = any(!is.na(d$opp_rank))}
ranks <-
  rbind(select(d, Date, team = opp,  rank = opp_rank),
        select(d, Date, team = home, rank = home_rank))
last_rank <-
  ranks %>%
  arrange(Date) %>%
  group_by(team) %>%
  summarize(last_rank = max(rank[Date == max(Date)]))

last_rank %>% filter(!is.na(last_rank)) %>% arrange(last_rank) 
```

# Perform Modeling

```{r prep data for modeling score differential}
dt_reg <-
  data.table(d) %>% 
  .[j = paste0("d_", teams) := lapply(teams, function(team) 1*(team == home) + (-1)*(team == opp))] %>%
  mutate(diff = Score - Opp.Score,
         days_since_start = Date - min(Date),
         days_scaled = as.integer(days_since_start) / max(as.integer(days_since_start)) + 1) %>%
  #select(-d_US_Under18_Team) %>%  # -d_Acadia, -d_TBD, 
  filter(!str_detect("TBD|US_Under18", home) & !grepl(opp, "TBD|US_Under18"))

filter(dt_reg, str_detect(home, "Minnesota_D")) %>% select(Date, Opp.Score, Score, opp, home, d_Minnesota_Duluth) %>% head()
filter(dt_reg, str_detect(opp,  "Minnesota_D")) %>% select(Date, Opp.Score, Score, opp, home, d_Minnesota_Duluth) %>% head()
```

## Analyze Goal Differential

### OLS

```{r ols}
inds <- str_subset(cn(dt_reg), "^d_")
# Select Wisconsin as the reference
inds <- setdiff(inds, "d_Wisconsin")

reg_fm <- as.formula(paste0("diff ~ ", paste(inds, collapse = " + ")))
reg <- lm(reg_fm, 
          weights = days_scaled^2,
          data = dt_reg)
reg_coeff <- tidy(reg) %>% mutate(team = gsub("^d_", "", term))

```

### Ordered Logistic

```{r ordered logstic}
# See this UCLA stats post with notes on using polr:
# https://stats.idre.ucla.edu/r/dae/ordinal-logistic-regression/
table(dt_reg$diff)

dt_ol <-
  dt_reg %>% 
  filter(diff %in% -6:6) %>%  # keep data within bounds that have at least some reasonable support
  mutate(fdiff = factor(diff, levels = sort(unique(diff))))

ol_fm <- reformulate(inds, "fdiff") #as.formula(paste0("fdiff ~ ", paste(inds, collapse = " + ")))
init <- reg_coeff %>% filter(!grepl("Intercept", term)) %>% pull(estimate)
ol <- polr(ol_fm, 
           method = "logistic", 
           data = dt_ol)
  
ol_coeff <- coef(ol) %>% data.frame(term = names(.), ol_est = .) %>% mutate(team = gsub("^d_", "", term))
```

### Compare Estimates

```{r compare ols and ordered reg}
comp <- merge(reg_coeff %>% rename(ols_est = estimate),
              ol_coeff,
              by = c("term", "team"))
ggplot(comp, aes(ols_est, ol_est)) +
  geom_point()
```

## Analyze Home and Opponent Scores Independently

The general modeling approach is analyzing each score in each game, to evaluate offense against defense in both directions. 

```{r develop data for analyzing individual scores}
dt_score <- 
  d %>% 
  data.table() %>% 
  mutate(game_id = 1:n(),
         days_since_start = Date - min(Date),
         days_scaled = as.integer(days_since_start) / max(as.integer(days_since_start)) + 1) %>%
  dplyr::select(game_id, Date, days_scaled, Opp.Score, Score, opp, home) %>% 
  melt(id.vars = c("Date", "days_scaled", "game_id", "opp", "home")) %>% 
  .[j = `:=`(o = ifelse(str_detect(variable, "Opp"),  opp, home),
             d = ifelse(str_detect(variable, "Opp"), home, opp))]
```

### Run Estimation Sensitivity

```{r lmer isn't working, eval = FALSE}
score_lmer <- 
  lmer(value ~ o + d + home + (1|game_id), 
       data = dt_score)

lmer_score_betas <- 
  coefficients(score_lmer)$game_id %>% 
  unique() %>% 
  melt() %>% 
  as.data.frame()
```

```{r multiple estimation methods}
score_ols <- 
  lm(value ~ o + d + home, 
     data = dt_score)

score_ols_wgt <- 
  lm(value ~ o + d + home, 
     weight = days_scaled^2,
     data = dt_score)

score_poisson <- 
  glm(value ~ o + d + home,
      family = "poisson",
      data = dt_score)

score_poisson_wgt <- 
  glm(value ~ o + d + home,
      family = "poisson",
      weights = days_scaled^2,
      data = dt_score)

score_betas <- 
  bind_rows(tidy(score_ols)         %>% mutate(method = "ols_unwgt"),
            tidy(score_ols_wgt)     %>% mutate(method = "ols_wgt"),
            tidy(score_poisson)     %>% mutate(method = "poisson_unwgt"),
            tidy(score_poisson_wgt) %>% mutate(method = "poisson_wgt")) %>% 
  mutate(first = str_extract(term, "^(.)"),
         type = case_when(first == "o" ~ "Offense",
                          first == "d" ~ "Defense",
                          first == "h" ~ "Home"))
```

### Compare and Select Among Estimates

```{r compare results across methods}
method_comp <-
  score_betas %>% 
  mutate(team = str_replace(term, "^(o|d|home)", "")) %>% 
  dplyr::select(team, estimate, type, method) %>% 
  filter(!is.na(type)) %>% 
  pivot_wider(names_from = method,
              values_from = estimate)

# Examine outliers
summary(method_comp)

method_comp_trim <-
  method_comp %>% 
  filter(ols_unwgt > -5,
         ols_wgt > -5,
         poisson_unwgt > -5,
         poisson_wgt > -5)

# method_comp %>% 
#   group_by(type) %>% 
#   summarize(rho = cor(cbind(ols_unwgt, ols_wgt)))#, poisson_unwgt, poisson_wgt)))

lapply(unique(method_comp$type),
       function(x) {
         method_comp_trim %>% 
           filter(type == x) %>% 
           with(cor(cbind(ols_unwgt, ols_wgt, poisson_unwgt, poisson_wgt)), use = "pairwise") %>% 
           round(3)
         })


plot_pairs <- function(dt) {
  options(warn = -1)
  pairs.panels(method_comp,
             smooth = TRUE,      # If TRUE, draws loess smooths
             scale = FALSE,      # If TRUE, scales the correlation text font
             density = TRUE,     # If TRUE, adds density plots and histograms
             ellipses = FALSE,    # If TRUE, draws ellipses
             method = "pearson", # Correlation method (also "spearman" or "kendall")
             pch = 21,           # pch symbol
             lm = FALSE,         # If TRUE, plots linear fit rather than the LOESS (smoothed) fit
             cor = TRUE,         # If TRUE, reports correlations
             jiggle = FALSE,     # If TRUE, data points are jittered
             factor = 2,         # Jittering factor
             hist.col = 4,       # Histograms color
             stars = TRUE,       # If TRUE, adds significance level with stars
             ci = TRUE)  
  options(warn = 0)
}

for (my_type in unique(method_comp$type)) {
  method_comp %>% 
    filter(type == my_type) %>% 
    dplyr::select(ols_unwgt, ols_wgt, poisson_unwgt, poisson_wgt) %>% 
    plot_pairs
}

# Choose preferred specification
score_betas_pref <- 
  score_betas %>% 
  filter(method == "poisson_wgt",
         estimate > -5)
```

### Examine Chosen Estimates

```{r visualize estimates}
score_betas_pref %>% 
  filter(!is.na(type)) %>% 
  ggplot(aes(x = estimate,
             color = type)) +
  geom_density()
```

```{r process groups of betas}
score_betas_pref_w <- 
  score_betas_pref %>%  
  filter(!is.na(type)) %>% 
  mutate(team = str_replace(term, "^(o|d|home)", "")) %>% 
  dplyr::select(team, estimate, type) %>% 
  pivot_wider(names_from = "type",
              values_from = "estimate") 
```

```{r typologize teams in the bracket}
tourney_teams_23 <- 
  c("Denver", "Cornell", "Western_Michigan", "Boston_University",
    "Minnesota", "Canisius", "Minnesota_State", "St_Cloud_State",
    "Michigan", "Colgate", "Michigan_Tech", "Penn_State",
    "Quinnipiac", "Merrimack", "Ohio_State", "Harvard")

stopifnot(all(tourney_teams_23 %in% score_betas_pref_w$team))

score_betas_pref_w <- 
  score_betas_pref_w %>% 
  mutate(tourney_team = 1*(team %in% tourney_teams_23))
```


```{r show scatter of best team offense and defense}
score_betas_pref_w %>% 
  ggplot(aes(x = -Defense,
             y = Offense,
             color = factor(tourney_team))) +
  geom_point()
```
```{r show just the tourney teams}
score_betas_pref_w %>% 
  filter(tourney_team == 1) %>% 
  ggplot(aes(x = -Defense,
             y = Offense)) +
  geom_point() +
  geom_text_repel(aes(label = team))
  
```

# Use Estimates for Selection

## Visualize Ranks from Goal Differential

```{r combine estimates with ranks}
ranks_b <-
  merge(x = filter(last_rank, !is.na(last_rank)),
        y = comp,
        by = "team",
        all.y = TRUE) %>%
  mutate(team = str_replace(team, "_", " ")) %>% 
  arrange(last_rank)
```

```{r visualize estimates vs ranks}
ggplot(ranks_b,
       aes(x = last_rank, 
           y = ol_est)) +
  geom_point() +
  geom_text(aes(label = team), size = 3) +
  scale_x_continuous(labels = ranks_b$last_rank, breaks = ranks_b$last_rank) +
  scale_y_continuous(limits = c(0, NA)) + 
  labs(x = "Team Rank", y = "Adjusted Goal Differential")
```

```{r print order of teams by strength per new data}
comp %>% 
  arrange(-ols_est) %>% 
  select(team, ols_est, ol_est) %>% 
  head(20)
```

## Visualize Ranks from Score-Based Analysis

```{r}
score_betas_pref_w %>% 
  filter(tourney_team == 1) %>% 
  mutate(Score = Offense + (-Defense)) %>% 
  arrange(-Score) %>% 
  transmute(Team = team, 
            Offense = round(Offense, 3), 
            Defense = round(Defense, 3), 
            Score   = round(Score, 3))
```


# Generate Monkey Bracket

```{r generate monkey bracket}
# These need to be provided in order on the bracket, i.e. the region, and then the seeding
bracket_teams <- sapply(c("NE", "E", "MW", "W"), function(x) paste0(x, c(1, 4, 3, 2))) %>% as.vector()

set.seed(as.integer(as.Date("2023-03-22")))
bracket_str <- sapply(bracket_teams, function(x) runif(1))

bracket_r <- bracket_str
for (round in 1:4){
  matchups <- length(bracket_r)/2
  bracket_r <- sapply(1:matchups, 
                      function(x) {
                        pos <- 2*x - 1
                        matchup <- bracket_r[pos:(pos+1)]
                        matchup[matchup == max(matchup)]
                      })
  print(paste0("Teams picked to win in Round ", round))
  print(names(bracket_r))
}

```


